Data Specialization for a Staged Lambda Calculus

### 2015-07-08 paper/splitting.tex
fleshed out splitting section

Top level programs in \lang\ are given by a term $\coltwo e A$.
Reduction of this program, via repeated $\stepsym \bbtwo$, will eventually yield a residual, which can be reduced no further.
We write this as:
$
e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q
$.

Splitting $e$ (via $\splittwosym$)
yields a pair of monostaged terms $\pipeS p l r$, where $p$ and $l.r$
contain all first- and second-stage computations in $e$ respectively.
This output is a multistage program in its own right.
It evaluates by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

The validity of the splitting translation requires that this simple 
evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
to the residual produced by direct evaluation, $q$.
This condition, called \emph{correctness},
is stated more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

@@

In the recursive case, we have 
$\typestwo [\cdot] e A$, $e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$, and $\splittwo e A p l r$.
To help this case, we introduce \ref{lem:twoCorrect}, whose proof will concern the rest of this section.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly. \qed

We prove \ref{lem:twoCorrect} by induction on the derivation of the step.
However, since second-stage steps can depend on first-stage steps,
we first cover the definition of splitting terms at \bbonem.

@@

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
The first component, the \emph{immediate value} $i$, is a monostage value holding
the result of all first-stage components of $v$.  The second
component, $q$, is a residual representing all second-stage
variable references of $v$.

To provide intuition about the intent of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`yhat`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:termSplit}.
Observe that partial value splitting does not lose any meaningful information (namely injection tags),
which is the validity requirement for a translation of value forms.

>>>>

Top level programs in \lang\ are given by a term $\coltwo e A$.
Reduction of this program, via repeated $\stepsym \bbtwo$, will eventually yield a residual, which can be reduced no further.
We write this as $e \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$.

Splitting $e$, via $\splittwosym$, yields $\pipeS p l r$.
This output is a multistage program in its own right.
It is evaluated by reducing $p$ to the value $b$ in the first pass,
then plugging that result in for $l$ to produce $[b/l]r$,
which is evaluated in the second pass.
Since execution of the first pass serves to generate input for
the second pass, we say $p$ is a {\em precomputation} that produces a
{\em boundary value} ($b$) for the {\em resumer} ($l.r$).

The correctness of the splitting translation requires that this simple 
evaluate-and-plug semantics produces the same answer as the staged semantics of the previous section.
That is, $[b/l]r$ (the \emph{applied resumer}) should be equivalent 
to the residual produced by direct evaluation, $q$.
This condition is stated more precisely as \ref{thm:allCorrect},
where ``$\reduce e v$'' indicates standard monostage reduction of the term $e$ to the value $v$,
and ``$\equiv$'' indicates a monostage equivalence, which we cover later.

@@

In the recursive case, where the evaluation takes at least one step, we have 
$e \stepsym \bbtwo e' \stepsym \bbtwo \cdots \stepsym \bbtwo \exv q$
as well as \mbox{$\typestwo [\cdot] e A$} 
and \mbox{$\splittwo e A p l r$}.
To help this case, we introduce \ref{lem:twoCorrect}, 
which essentially states that any single step is correct,
and whose proof will concern the rest of this section.
By preservation and totality of splitting, we know $\typestwo [\cdot] {e'} A$ and $\splittwo {e'} A {p'} {l'} {r'}$.
The inductive hypothesis yields $\reduce {p'} {b'}$ and $[b'/l']r' \equiv q$.
From here, \ref{lem:twoCorrect} then gives $\reduce p b$ and $[b/l]r \equiv [b'/l']r'$,
and we derive $[b/l]r \equiv [b'/l']r' \equiv q$ directly.

This concludes the proof of program correctness, modulo \ref{lem:twoCorrect}.
However in order to prove that lemma, 
we will need to state analogous version for steps at \bbonem,
since the various kinds multistage term in \lang\ are mutually dependent.
Thus, this section proceeds by covering the definition of splitting at \bbonem, 
starting with the value form at that world.

@@

Partial value splitting (given by the relation $\vsplito$) converts a 
partial value $v$ into a split value $\mval i q$.
Since partial values, by definition, have no remaining work in the first pass
and since transfer of information between the stages occurs in the first pass of evalution,
we know that the second stage components of $v$ can no longer depend on it's first stage components.
Analogously, this must hold of the output of splitting, 
which is why $q$---unlike the resumers of world \bbtwo\ term splitting---is not open on a variable.

Thus instead of a pipeline relationship, $i$ and $q$ bear more of a parallel relationship.
That is, they embody the complementary portions of $v$ that are relevant to each stage.
For this reason we call $i$ the \emph{immediate value} and $q$ the residual.

To provide intuition about the behavior of partial value splitting, consider the
following partial value:
\begin{lstlisting}
1` (next{`2`y`1`}, (`3`gr{`1`injL 7`3`}`1`, next{`2`y`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
splitting first redacts all second-stage (blue) parts, along with the
surrounding \texttt{next} annotations. 
The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), (3`gr{`1`injL 7`3`}`,()))`
\end{lstlisting}
Finally, partial value splitting drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), (injL 7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second-stage computations) partial value splitting redacts all \texttt{gr} blocks
(replacing them with unit) and \texttt{next} annotations:
\begin{lstlisting}
2`(y, ((),y))`
\end{lstlisting}
A precise definition of the partial value splitting relation is given in
\ref{fig:termSplit}.

In some regards, the exact formulation of partial value splitting is arbitrary.
For instance, we chose to replace ``holes'' with unit values,
but in fact we could have used any value there and it would make no difference in the end.
There are however, at least some parts of the definition that are not arbitrary.
Importantly, partial value splitting must not lose any meaningful information,
such as injection tags.
